### 2018-09-04 Manuscripts/cog-psych/revision/cogpsych.tex
Primary revisions to manuscript: copytweak-rep in section 8.5, unsup cat, contrast in thomas98, some mention of machine learning generative models. In other news, updated and more or less finalised fits with the category [0] bug fixed in Simulation.py. TODO: Update the goldstate.

\begin{table} \centering
\caption{Results of model-fitting to the combined datasets from Experiments 1
and 2. Note that smaller AIC values correspond to better model fits (adjusted
for number of parameters)}
\label{table:global-model-fits}
\begin{tabular}{ l l l l} \\ \textbf{PACKER} & \textbf{Copy \& Tweak} &
\textbf{Representativeness} & \textbf{Hierarchical Bayesian} \\ \hline $AIC =
  9095$ & $AIC = 9842$ & $AIC = 8878$ &  $AIC =
9912$ \\ $L = -4545$ & $L = -4919$ & $L = -4435$ & $L = -4952$ \\ $c = 0.483$ &
                                                                                $c = 3.197$ & $\kappa = 11.232$ &
$\kappa < 0.001$ \\ $\theta_c = 3.16$ & $\theta_c = 0$ (fixed) & $\nu = 2.844$ & $\nu = 5.597$ \\
$\theta_t = 3.494$ & $\theta_t = 2.972$ & $\lambda = 1.917$& $\lambda
                                                                       = 0.055$
  \\ & &$\theta = 5.885 $& $\theta = 3.175$
\\
\end{tabular}
\end{table}

@@

\begin{figure}
    \begin{center} \inputpgf{figs/}{packer-loglike.pgf}
      \caption{PACKER's fit as a function of its prioritization of
        within-category and between-category similarity (using the
        $\theta_c$ parameter) . To facilitate comparison, PACKER's other
        parameters ($c$, $\theta_t$) were set to the best fitting values
        obtained for copy-and-tweak in Table \ref{table:global-model-fits}. }
    \label{fig:packer-loglike}
    \end{center}
\end{figure}

@@

\section{General Discussion}

The creative generation of concepts is an immensely intriguing topic, but it is
also complex and difficult to study in a traditional laboratory setting.
Creative generation is most typically studied using a creative cognition
approach \citep{finke1992creative,smith1995creative}, where creative products
\citep[such as drawings of alien plants and animals;][]{ward1994structured} are
analyzed to obtain insight into the processes and representations involved.
While this approach has yielded a great deal of information about the nature of
creative generation, the commonly employed response modalities (e.g., hand-drawn
pictures) make it difficult to develop computational models capable of formally
evaluating the proposed representations and processes. However, the generation
of novel concepts is not altogether different from other behaviors studied in
cognitive psychology: In particular, it can be viewed as an form of inference
using one's knowledge about existing categories
\citep{kemp2014taxonomy,kurtz2015human}. Taking advantage of this insight, in
this paper we investigated the generation of new categories using the empirical
and computational toolkit from the field of categorization.

While the bulk of prior research on the topic has focused on the classic finding
that generated concepts tend to be distributionally similar to known concepts,
there has been little work addressing the role of contrast in creative
generation: How is it that people are able to create something {\em different}
from what is already known? We developed a novel, exemplar-based model, PACKER,
which formally specifies the role of contrast in generation. The model proposes
that categories are represented as exemplars in a multidimensional psychological
space, and generation is constrained both by within-category and
between-category similarity: Exemplars belonging to the same category should be
similar to one another, and exemplars belonging to different categories should
not be similar to one another.

In addition to an analysis of published data \citep[][Experiment
3]{jern2013probabilistic}, we reported two experiments demonstrating systematic
effects of category contrast in creative generation. Members of
participant-generated categories tended to be highly dissimilar from members of
previously-learned categories, and were usually more similar to one another than
to members of other categories. We also observed broad interdependence between
the distributional structure (feature variance, correlation) and physical
instantiation (location within the stimulus space) of generated categories: In
Experiment 2, we found that the unoccupied regions of the domain influenced the
distributional structure of categories, and in both experiments we observed that
participants distributed their generated categories to increase contrast with
what was already known.

We conducted simulations comparing PACKER's account of our results to the other
major proposals for category generation: a ``copy-and-tweak'' model (realized as
a variant of PACKER with no sensitivity to category contrast), and a
hierarchical Bayesian model designed to explain the classic distributional
similarity effect. In all simulations, we found that PACKER's sensitivity to
contrast captured a previously unexplained and unexplored aspect of human
category generation. Further, by measuring PACKER's fit as a function of its
prioritization of within- and between-category similarity, we observed that
considering either constraint exclusively results in a relatively low-quality
account. Instead, PACKER's best results were obtained when both constraints are
considered, indicating that human learners do not generate novel concepts
exclusively on the basis of within-category similarity or between
class-contrast. This finding mirrors our behavioral results and demonstrates
that both constraints influence creative generation.

@@

\subsection{Disentangling Exemplar dissimilarity and Representativeness in Categorization}

\sxlnote{I'm struggling a little with this section! That said, I think this may
  also be a good point to address Charles' concern re: the possible conditions
  under which contrast are important.} 
Although both PACKER and the
representativeness model perform better than the other models, they appear to be
capturing fundamentally different aspects of contrast in categorization.
Specifically, while the representativeness model excelled in predicting our
category generation data, PACKER was the best-fitting model in accounting for
participant error in category learning. The source of this discrepancy is still
not yet clear -- it could be an indication of a fundamental difference between
the processes involved in generation versus learning where exemplar
dissimilarity is a larger factor in category generation while representativeness
exerts more influence in category learning.

In this paper, we have developed two novel models of category generation, each
involving a different conceptualization of contrast.  However, there is no theoretical constraint that dictates the It is also possible to treat the similarity

A thorough investigation to disentangle the two contrast mechanisms here is
beyond the scope of the current article. However, as a quick proof of concept it
is useful to check if the addition of a representativeness mechanism to an
exemplar model would yield any benefit at all.

Xian: Please
write about the two hypotheses and that followup work needs to look into why the
Bayesian representativeness model doesn't do as well at explaining category
learning ease, but does better at category generation.

MAKE SURE WE CITE AND DESCRIBE \citet{thomas98}! \sxlnote{I moved the
  description and citation of thomas98 to the introduction for justification of
  why we think categories should share similar distributions.}
\subsection{Limitations and Future Directions} Although we have demonstrated
that contrast plays a role in categorization, our results were equivocal as to
whether contrast is governed by exemplar dissimilarity or representativeness.

@@

Finally, although an overall goal of this work is to advance the study of
creative thought by explaining generation through empirical and quantitative
approaches from research in categorization, these approaches are not easily
aligned with many of the alternative theoretical viewpoints in the broader study
of creativity \citep[for a review see][]{kozbelt2010theories}, such as those
based on free association \citep{mednick1962associative} and conceptual
combination \citep{estes2002emergence,murphy1988comprehending}. Instead, we
reduced our focus by studying a highly complex behavior (creative generation) as
it applies within a well-established domain (artificial category learning). We
hope that future work incorporates and highlights the importance of contrast
into theories of creativity.

>>>>

\begin{table} \centering
\caption{Results of model-fitting to the combined datasets from Experiments 1
and 2. Note that smaller AIC values correspond to better model fits (adjusted
for number of parameters)}
\label{table:global-model-fits}
\begin{tabular}{ l l l l} \\ \textbf{PACKER} & \textbf{Copy \& Tweak} &
\textbf{Representativeness} & \textbf{Hierarchical Bayesian} \\ \hline $AIC =
  9069$ & $AIC = 9813$ & $AIC = 8783$ &  $AIC =
9881$ \\ $L = -4531$ & $L = -4905$ & $L = -4388$ & $L = -4937$ \\ $c = 0.51$ &
                                                                                $c = 3.22$ & $\kappa = 12.23$ &
$\kappa < 0.001$ \\ $\theta_c = 3.09$ & $\theta_c = 0$ (fixed) & $\nu = 1.00$ & $\nu = 5.44$ \\
$\theta_t = 3.47$ & $\theta_t = 3.00$ & $\lambda = 7.04$& $\lambda
                                                                       = 0.06$
  \\ & &$\theta = 10.21 $& $\theta = 3.09$
\\
\end{tabular}
\end{table}

@@

\begin{figure}
    \begin{center} 
      \includegraphics[width=\textwidth]{figs/packer-loglike.pdf}
      \caption{PACKER's fit as a function of its prioritization of
        within-category and between-category similarity (using the $\theta_c$
        parameter) . To facilitate comparison, PACKER's other parameters ($c$,
        $\theta_t$) were set to the best fitting values obtained for
        copy-and-tweak in Table \ref{table:global-model-fits}. The black diamond
        marker indicates the log-likelihood for the point where
        $\theta_c=\theta_t$. }
    \label{fig:packer-loglike}
    \end{center}
\end{figure}

@@

\section{General Discussion}
The sensory impression of every stimulus and event is unique. Grouping distinct
patterns of sensory information into categories is a fundamental task solved by
the mind. Most work has focused on how people learn new categories that are
provided to them through unlabeled, partially labeled, or fully-labeled
examples. How were these categories first determined? Some natural categories
are likely to be the result of regularities in the dynamics of our environment.
But, these are only a subset of the categories that people learn. Other
categories, such as tools and ideas, were generated by people over time. What
basic principles underlie how people generate categories?

% The creative generation of concepts is an immensely intriguing topic, but it is
% also complex and difficult to study in a traditional laboratory setting.
% Creative generation is most typically studied using a creative cognition
% approach \citep{finke1992creative,smith1995creative}, where creative products
% \citep[such as drawings of alien plants and animals;][]{ward1994structured} are
% analyzed to obtain insight into the processes and representations involved.
% While this approach has yielded a great deal of information about the nature of
% creative generation, the commonly employed response modalities (e.g., hand-drawn
% pictures) make it difficult to develop computational models capable of formally
% evaluating the proposed representations and processes. However, the generation
% of novel concepts is not altogether different from other behaviors studied in
% cognitive psychology: In particular, it can be viewed as an form of inference
% using one's knowledge about existing categories
% \citep{kemp2014taxonomy,kurtz2015human}. Taking advantage of this insight, in
% this paper we investigated the generation of new categories using the empirical
% and computational toolkit from the field of categorization.

While the bulk of prior research on the topic has focused on the classic finding
that generated concepts tend to be distributionally similar to known concepts,
there has been little work addressing the role of contrast in creative
generation: How is it that people are able to create something {\em different}
from what is already known? \jlanote{Update this next section:}
\sxlnote{Updated} We developed two novel models, each incorporating a different
conceptualization of category contrast. One model is PACKER, which is an
exemplar-based model that formally specifies the role of contrast in generation
as exemplar dissimilarity. Specifically, the model proposes that categories are
represented as exemplars in a multidimensional psychological space, and
generation is constrained both by within-category and between-category
similarity: Exemplars belonging to the same category should be similar to one
another, and exemplars belonging to different categories should not be similar
to one another. The second model is a novel hierarchical Bayesian model with a
representativeness mechanism. This model generates exemplars that are more
representative of (i.e., has greater relative evidence from) the novel category
compared to the learned category.

We reported two experiments demonstrating systematic effects of category
contrast in creative generation. Members of participant-generated categories
tended to be highly dissimilar from members of previously-learned categories,
and were usually more similar to one another than to members of other
categories. We also observed broad interdependence between the distributional
structure (feature variance, correlation) and physical instantiation (location
within the stimulus space) of generated categories: In Experiment 2, we found
that the unoccupied regions of the domain influenced the distributional
structure of categories, and in both experiments we observed that participants
distributed their generated categories to increase contrast with what was
already known.

We conducted simulations comparing the contrast models' account of our results
to the classical proposals for category generation: a ``copy-and-tweak'' model
(realized as a variant of PACKER with no sensitivity to category contrast), and
a hierarchical Bayesian model designed to explain the classic distributional
similarity effect. In all simulations, we found that the contrast models captured a previously unexplained and unexplored aspect of human
category generation. In particular, by measuring PACKER's fit as a function of its
prioritization of within- and between-category similarity, we observed that
considering either constraint exclusively results in a relatively low-quality
account. Instead, PACKER's best results were obtained when both constraints are
considered, indicating that human learners do not generate novel concepts
exclusively on the basis of within-category similarity or between
class-contrast. This finding mirrors our behavioral results and demonstrates
that both constraints influence creative generation when explained with an
exemplar model. Ultimately, our category generation data was best explained by
the representativeness model. The exact nature of its superiority over the other
models is still not yet clear. However, as we explore in a later section, it is
likely that the representativeness mechanism itself (i.e., independent of the
hierarchical Bayesian foundation) captures a core aspect of both category
generation and category learning.
\jlanote{Above needs some updating still. Also, do we want an outline of general
discussion paragraph?}
\sxlnote{Updated. I tried writing an outline, but found that points got a little
  repetitive after a bit, so I ditched that idea...}

\subsection{Novel Analyses Demonstrating Contrast Effects in Prior Work}

It is possible, however, that this data can be explained by a process that does
not involve contrast. Specifically, since participants were trained on exemplars
that share the same hue within a category, it is plausible that they generate
exemplars that all share a common hue. If this is the case, given a $6/8 = .75$
probability that a hue distinct from the experimenter-defined categories is
selected, there is a corresponding $.75$ probability (or about 17 out of 22
participants) that the generated category will have no exemplars that share a
hue with any of the experimenter-defined categories. In addition, that there is
a $2/8 = .25$ probability (about 6 out of 22 participants) that the generated
category will have all six exemplars that share a hue with any of the
experimenter-defined categories. This prediction of regularity across categories
is much closer to the data than the prediction from a Binomial model and does
not require any mechanism of category contrast.

\jlanote{Xian: Please do a Chi-Squared test between what is expected based on
  the logic above and the observed frequencies mentioned below} \sxlnote{I
  didn't do a chi-sq test because what appears below is not the observed
  frequencies that correspond to the logic above, but is instead an indication
  of whether or not it is reasonable to assume the logic above (and our argument
  is that it is not reasonable.)}

@@

The current work also suggests exciting directions for related investigations
into unsupervised categorization. While the unsupervised categorization
literature is primarily interested in the generation of categories within a set
of observed exemplars where no prior categories are learned, our category
generation studies are focused on the production of novel category exemplars
themselves. Despite this distinction, both types of categorization research are
ultimately interested in the question of how categories can be formed. The results
of our current work indicate that the formation of categories is constrained by
some measure of contrast. To our knowledge, this explicit investigation of
category contrast has not yet been attempted in the unsupervised categorization
literature.

Interestingly, the unidimensional sort bias that is commonly observed in
unsupervised categorization
\citep{imai1965discriminability,milton2004influence,ahn1992two} was not
consistently observed in our tasks. To clarify, the unidimensional sort bias is
the observation that categories are typically constructed solely by focusing on
a single dimension or feature. In our experiments we only observed this bias in
the row condition of Experiment 1, whereas in other conditions participants
generated categories that typically varied along both dimensions. This may be
unsurprising at face value, since participants in conditions other than the row
condition were trained on a prior category that varied on both dimensions.
However, these results suggest that the unidimensional sort bias is not
universal to all types of category construction -- or more specifically, the
unidimensional bias may not apply when participants need to generate categories
with entirely new exemplars.

\subsection{Exemplar dissimilarity and Representativeness in Categorization} 
\sxlnote{Hmmm.. ok after quite a few revisions I am still not quite satisfied
  with this section. It doesn't feel quite ...tight, but I think it's time for
  me to say stop and pass it on.}
Although both PACKER and the representativeness model perform better than the
classical models, they appear to be capturing fundamentally different aspects of
contrast in categorization. Specifically, while the representativeness model
excelled in predicting our category generation data, PACKER was the best-fitting
model in accounting for participant error in category learning. Given the
history of their development, this is perhaps not entirely surprising. PACKER
is implemented as an extension of the GCM, which was originally developed for (and
has been highly successful in) category learning \citep{nosofsky1986attention,
  nosofsky1994comparing}. Conversely, the representativeness mechanism was
conceived as a heuristic describing the relative extent to which an individual
observation is generated from some unobserved population
\citep{kahneman1972subjective,abbott16}. This generative nature of the
representativeness mechanism intuitively lends itself to the prediction of
category generation data.

While the contrast models were born quite naturally out of the corresponding
classical models, it is worth considering if similar benefits in prediction
would be seen if the contrast mechanisms were applied to different classical
models. Although there is no clear way to adapt the exemplar dissimilarity
mechanism to the current hierarchical Bayesian model, it is a fairly
straightforward process applying the representativeness mechanism to an exemplar
model. Specifically, by treating the similarity measure (Equation
\ref{eq:similarity}) in the current implementation of copy-and-tweak as a
density estimate for the representativeness mechanism (Equation
\ref{eq:representativeness}), we gain a new model of representativeness that is
based on exemplar similarity instead of a multivariate Gaussian
likelihood.\footnote{An analogous implementation of the representativeness
  mechanism to PACKER results in both $\theta_c$ and $\theta_t$ adding to
  constant value that is independent of the similarity to contrast and target
  exemplars respectively. Consequently, it becomes equivalent to a
  copy-and-tweak model with representativeness.}

By subjecting this preliminary model to the same analysis as the other models,
we find that it substantially outperforms its classical counterpart, but not to
the extent of PACKER or the representativeness model. Specifically, fitting this
model to the category generation data yielded $L = -4815$ and $AIC=9633$ for the
entire set of data from Experiments 1 and 2, and $L = -3474$, $AIC=6953$ for
data excluding the first trials. When used to predict the error rates of
participants in the category learning task (Experiment 3), the performance of
the copy-and-tweak with representativeness model correlated with participant
error with $r=.56, p<.001$. These results suggest that the predictive advantages
to including the representativeness mechanism is not limited to a hierarchical
Bayesian model, but can also be found when applied to an exemplar model.
Although its performance is not as strong as either of the fully-developed
contrast models, there is a benefit to the representativeness mechanism that is
independent from any interaction with a hierarchical Bayesian framework.

Ultimately, the source of the discrepancy between the performance of the two
contrast mechanisms in category generation versus category learning is still not
entirely clear. Our results could be an indication of a fundamental difference
between the processes involved where exemplar dissimilarity is a larger factor
in category generation while representativeness exerts more influence in
category learning. While a thorough investigation is beyond the scope of the
current article, the preliminary results of this section indicate that further
work to disentagle these two mechanisms would be promising. In the following
section, we explore some other limitations of the current work and provide more
recommendations for future research.

\subsection{Limitations and Future Directions} Although we have demonstrated
that contrast plays a role in categorization, our results were equivocal as to
whether contrast is governed by exemplar dissimilarity or representativeness.

@@

Finally, although an overall goal of this work is to advance the study of
creative thought by explaining generation through empirical and quantitative
approaches from research in categorization, these approaches are not easily
aligned with many of the alternative theoretical viewpoints in the broader study
of creativity \citep[for a review see][]{kozbelt2010theories}, such as those
based on free association \citep{mednick1962associative} and conceptual
combination \citep{estes2002emergence,murphy1988comprehending}. In addition, the
machine learning literature has taken very interesting approaches using neural
networks to address the overall problem of creative generation
\citep[e.g.,][]{goodfellow2014gan,ho2016gail,kingma2016,chen2016infogan}. In
contrast to these varied investigations, we reduced our focus by studying a
highly complex behavior (creative generation) as it applies within a
well-established domain (artificial category learning). We hope that future work
incorporates and highlights the importance of contrast into theories of
creativity.

### 2018-09-04 Manuscripts/cog-psych/revision/methods_results_e3.tex
Primary revisions to manuscript: copytweak-rep in section 8.5, unsup cat, contrast in thomas98, some mention of machine learning generative models. In other news, updated and more or less finalised fits with the category [0] bug fixed in Simulation.py. TODO: Update the goldstate.

Although category generation is an intriguing aspect of human cognition, it has
played a relatively limited role in the broader categorization literature. To
establish that contrast plays a broader role in categorization, we conducted a
traditional category learning study. In it, participants learned two categories,
where the categories learned by each participant in Experiment 3 was yoked to
the Alpha shown and Beta generated by each participant in Experiment 2. Our goal
is to investigate whether model performance (in generating categories) is
consistent with human performance (in learning those categories). Specifically, if contrast plays a broader role in categorization, then Beta categories rated as "better" according
to the contrast models should be easier for participants to learn.
%\footnote{It is possible to adjust the present models of category generation such that they can make category learning predictions. However, such an adjustment makes the strong assumption that the very same processes governing category generation also apply to category learning. Taking this approach would fall outside the scope of the current investigation.} 
\subsection{Participants, Materials, and Procedure}

@@

Overall accuracy of the participants was high, with a mean error rate of $.19$
($SD = .19$). Aggregated error rates for each block are presented in Figure
\ref{fig:learningcurve}.

@@

The greatest correlations were observed when fitting with the contrast models.
Specifically, PACKER correlated with human performance with
$r = .65, \ 95\% CI$\footnote{Confidence intervals were obtained using a bootstrap
  method with 1000 bootstrap samples.}$ \ [.47,.78]$, and the representativeness
model correlated with human performance with $r = .56, \ 95\%  CI \ [.38, .72]$.
Copy-and-tweak yielded a correlation of $r = .45, 95\% CI \ [.29, .61]$ and the
hierarchical Bayesian model yielded a similar correlation of
$r = .46, 95\% CI \ [.30, .61]$. These results are graphically presented in Figure
\ref{fig:perror_corr}.

>>>>

Although category generation is an intriguing aspect of human cognition, it has
played a relatively limited role in the broader categorization literature. To
establish that contrast plays a broader role in categorization, we conducted a
traditional category learning study. In it, participants learned two categories,
where the categories learned by each participant in Experiment 3 was yoked to
the Alpha shown and Beta generated by each participant in Experiment 2. Our goal
is to investigate whether model performance (in generating categories) is
consistent with human performance (in learning those categories).
%\footnote{It is
  % possible to adjust the present models of category generation such that they
  % can make category learning predictions. However, such an adjustment makes the
  % strong assumption that the very same processes governing category generation
  % also apply to category learning. Taking this approach would fall outside the
  % scope of the current investigation.} 
Specifically, if contrast plays a
broader role in categorization, then Beta categories rated as "better" according
to the contrast models should be easier for participants to learn.

\subsection{Participants, Materials, and Procedure}

@@

Overall accuracy of the participants was high, with a mean error rate of $.19$
($SD = .19$). Error rates for each block are presented in Figure
\ref{fig:learningcurve}.

@@

The greatest correlations were observed when fitting with the contrast models.
Specifically, PACKER correlated with human performance with
$r = .65, 95\% CI$\footnote{Confidence intervals were obtained using a bootstrap
  method with 1000 bootstrap samples.}$ [.48,.78]$, and the representativeness
model correlated with human performance with $r = .59, 95\% CI [.40, .73]$.
Copy-and-tweak yielded a correlation of $r = .45, 95\% CI [.28, .62]$ and the
hierarchical Bayesian model yielded a similar correlation of
$r = .46, 95\% CI [.30, .62]$. These results are graphically presented in Figure
\ref{fig:perror_corr}.
